---
title: "Halloween Mini Project Class 9"
author: "Matthew White"
format: pdf

---

```{r}
candy <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names=1)

```

```{r}
head(candy)
```

>Q1: How many different candy types in dataset?  85
>Q2: How many fruity candy types are in dataset?  38

```{r}
nrow(candy)
table(candy["fruity"])
sum(candy["fruity"])
```

>Q3: What is your favorite candy in the dataset and its `winpercent` value? 50.35
>Q4: What is the `winpercent` value for "Kit Kat"?  76.76
>Q5: What is the `winpercent` value for "Tootsie Roll Snack Bars"? 49.65

```{r}
#rownames(candy)
candy["Almond Joy", "winpercent"]
candy["Almond Joy",]$winpercent

candy["Kit Kat", "winpercent"]
candy["Tootsie Roll Snack Bars", "winpercent"]

#Could also do this all at once in pipe syntax, once load dplyr package
#The %in% function is asking which of the following info is within the previous vector/data frame
library(dplyr)
candy %>% 
    filter(rownames(candy) %in% c("Almond Joy", "Kit Kat", "Tootsie Roll Snack Bars")) %>%
    select(winpercent)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?  A: winpercent is on a different scale to the rest of the columns. 

>Q7. What do you think a zero and one represent for the candy$chocolate column? 
Whether a candy has chocolate or not. 



```{r}
library("skimr")
skim(candy)


```

>Q8. Plot a histogram of winpercent values
      see code
>Q9. Is the distribution of winpercent values symmetrical?
      distribution is not symmetrical
>Q10. Is the center of the distribution above or below 50%?
      center is below 50%, at 47%
>Q11. On average is chocolate candy higher or lower ranked than fruit candy?
      Chocoalte is higher on avg
>Q12. Is this difference statistically significant?

```{r}
library("ggplot2")

#Q8, 9, 10
ggplot(candy, aes(winpercent)) + geom_histogram() + geom_vline(aes(xintercept = median(winpercent)), color = "red", linetype = "dashed")

median(candy$winpercent)

#Q11, 12
chocolate_avg_rank <- candy %>%
    filter(chocolate == 1) %>%
    select(winpercent)

fruity_avg_rank <- candy %>%
    filter(fruity == 1) %>%
    select(winpercent)

mean(chocolate_avg_rank$winpercent) > mean(fruity_avg_rank$winpercent)

t.test(chocolate_avg_rank$winpercent, fruity_avg_rank$winpercent)


```


>Q13. What are the five least liked candy types in this set?
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

>Q14. What are the top 5 all time favorite candy types out of this set?
Reeses PB Cup, Reeses Mini, Twix, Kit Kat, Snickers

```{r}
#tell the candy number (row number) with the lowest to highest winpercent order
inds <- order(candy$winpercent)
head(candy[inds,], 5)

(tail(candy[inds,], 5))

#can also use decreasing = T argument inside order function or rev() function to reverse the indeces

```

>Q15. Make a first barplot of candy ranking based on winpercent values.

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

> Q17. What is the worst ranked chocolate candy?
Sixlets
> Q18. What is the best ranked fruity candy?
Starburst

```{r}
ggplot(candy, aes(winpercent, rownames(candy))) + geom_col()

ggplot(candy, aes(winpercent, reorder(rownames(candy), winpercent))) + geom_col()

#rep function replicates first argument for specified number of times
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy, aes(winpercent, reorder(rownames(candy), winpercent))) + geom_col(fill = my_cols)
```



>Extra: Color favorite candy by favorite color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols[as.logical(rownames(candy) == "Almond Joy")] <- "blue"

# ?Could use this syntax: candy[,"Almond Joy"] if the candy names were still themselves a column, rather than rownames? 

ggplot(candy, aes(winpercent, reorder(rownames(candy), winpercent))) + geom_col(fill = my_cols)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
      Reese's mini have highest winpercent at lowest pricepercent
      
>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
      Least popular of the 5 most expensive candies is Nik L Nip 
      
```{r}
library(ggrepel)

ggplot(candy, aes(winpercent, pricepercent, label = rownames(candy))) + geom_point(col = my_cols) + geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

#c(11,12) gives us only the price and winpercent columns of the ordered rows we ask for. 
candy_price_order <- order(candy$pricepercent, decreasing = TRUE)
head(candy[candy_price_order,c(11,12)], 5)
```


>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy, aes(pricepercent, rownames(candy))) + geom_col()

#order x-axis by value
ggplot(candy, aes(pricepercent, reorder(rownames(candy), pricepercent))) + geom_col()

#make a dot/lollipop chart
ggplot(candy, aes(pricepercent, reorder(rownames(candy), pricepercent))) + geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") + geom_point()

```


##Explore the correlation structure 

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
      Fruity and chocolate candies are anti-correlated. Bar and pluribus candies too, understandably.

>Q23. Similarly, what two variables are most positively correlated?
      Chocolate is highly correlated with winpercent!

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)

#I dig this plot
```


>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? 
      Fruity, hard, and pluribus variables. These variables are highly anticorrelated with other variables such as chocolate and bar and winpercent that also have high magnitudes in PC1 but in the opposite direction.   

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)

#recall that x column contains the pca coordinates
plot(pca$x[,1], pca$x[,2], col = my_cols, pch =16)

#tell where each type of candy lies on the pca pc1 column
pca$rotation[,1]
#barplot of the pca1 coordinates for our candies, par function sets the margians
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")

```

```{r}
#putting pca results for pc1,2, and 3 into our candy data frame to help with ggplot functionality
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p

#use the ggrepel function
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```




